Aligning marine species range data to better serve science and conservation

### 2016-03-04 data_explore/data_explore.Rmd
updating refs, numbers etc in manuscript

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           progress = TRUE)

@@

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### this file created in data_prep
spp_list <- read.csv(file.path(dir_data, 'spp_list_w_area_trimmed.csv'), 
                     stringsAsFactors = FALSE)

>>>>

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           progress = TRUE,
                           col_types = 'cddddc')

@@

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### this file created in data_prep
spp_list <- read_csv(file.path(dir_data, 'spp_list_w_area_trimmed.csv'))

### 2016-03-04 data_explore/data_prep.Rmd
updating refs, numbers etc in manuscript

library(readr)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

@@

if(basename(getwd()) != 'data_explore') setwd('~/github/IUCN-AquaMaps/data_explore')
# source('data_explore_fxns.R')
### When knitting this, it automatically sets WD to be this directory...
### the 'setwd()' is there for running by chunk
```

@@

# Data files
#iucn spp cells WITHOUT extinct locations of IUCN subpopulations
message('Building IUCN species to cell table.  This might take a few minutes.')
iucn_map_files <- file.path(dir_anx, 'iucn_intersections', list.files(file.path(dir_anx, 'iucn_intersections')))
iucn_spp_cells <- lapply(iucn_map_files, data.table::fread) %>%
  bind_rows()
iucn_spp_cells <- iucn_spp_cells %>%
  rename(iucn_sid = id_no, loiczid = LOICZID) %>%
  mutate(iucn_sid = as.integer(iucn_sid),
         presence = as.integer(presence),
         loiczid  = as.integer(loiczid)) %>%
  filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
  filter(presence != 5)
### length(unique(iucn_spp_cells$iucn_sid)) results in 2013 species found...
### some species with no cells reported.  2249 listed in the am v iucn file.
write_csv(iucn_spp_cells, file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'))

message('Loading AquaMaps species to cell table.  This might take a few minutes.')
am_spp_cells <- fread(file.path(dir_data_remote, 'aquamaps/v2015/csv/hcaf_sp_native_trunc.csv'), 
                         showProgress = TRUE)
am_spp_cells <- am_spp_cells %>%
  rename(am_sid = speciesid, prob = probability) %>%
  filter(am_sid %in% spp_am_v_iucn$am_sid)
### length(unique(am_spp_cells$am_sid)) results in 2283 species found - great!
write_csv(am_spp_cells, file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'))

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload <- FALSE
if (!file.exists(data_file) | reload) {
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))

@@

spp_list <- spp_list %>%
  mutate(sm_perc = ifelse(is.na(sm_perc), 0, sm_perc)) %>%
  filter(!area_iucn == 0)
  ### will ditch the ones where no IUCN polygon exists - or are these species
  ### whose ranges only occur in cells beyond the LOICZID grid?

>>>>

library(raster)
library(readr)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

@@

if(basename(getwd()) != 'data_explore') setwd('~/github/IUCN-AquaMaps/data_explore')
source('data_explore_fxns.R')
### When knitting this, it automatically sets WD to be this directory...
### the 'setwd()' is there for running by chunk
```

@@

# Data files

### This file is generated in globalprep/SPP_ICO:
iucn_spp_cells_all <- read_csv(file.path(dir_data_remote, 'iucn_spp/iucn_cells_v2016.csv'),
                            col_types = 'cddddc')

iucn_spp_cells <- iucn_spp_cells_all %>%
  rename(iucn_sid = id_no) %>%
  filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
  filter(presence != 5)
### length(unique(iucn_spp_cells$iucn_sid)) results in 2279 species found...
### some species with no cells reported.  2279 listed in the am v iucn file.
write_csv(iucn_spp_cells, file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'))

message('Loading AquaMaps species to cell table.  This might take a few minutes.')
am_spp_cells_all <- fread(file.path(dir_data_remote, 'aquamaps/v2015/csv/hcaf_sp_native_trunc.csv'), 
                         showProgress = TRUE)
am_spp_cells <- am_spp_cells_all %>%
  rename(am_sid = speciesid, prob = probability) %>%
  filter(am_sid %in% spp_am_v_iucn$am_sid)
### length(unique(am_spp_cells$am_sid)) results in 2314 species found - great!
write_csv(am_spp_cells, file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'))

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload <- FALSE
if (!file.exists(data_file) | reload) {

  if(!exists('am_spp_cells'))
    am_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'))
  if(!exists('iucn_spp_cells'))
    iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'),
                               col_types = 'cddddc')
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))

@@

spp_list <- spp_list %>%
  mutate(sm_perc = ifelse(is.na(sm_perc), 0, sm_perc)) %>%
  filter(!area_iucn == 0)
  ### will ditch the ones with no IUCN area - formerly caused by DD species
  ### being filtered out.  Now, none suffer this consequence.

### 2016-03-04 ms_1_abst_sig.Rmd
updating refs, numbers etc in manuscript

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps and the International Union for Conservation of Nature (IUCN), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(nrow(spp_all), big.mark = ',')` species, mostly from AquaMaps, with only `r format(nrow(spp_map_pairs), big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index biodiversity goal and a global analysis of gaps in coverage of marine protected areas - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on which dataset was used. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___
<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

>>>>

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps and the International Union for Conservation of Nature (IUCN), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(n_spp_all, big.mark = ',')` species, mostly from AquaMaps, with only `r format(n_spp_both, big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index biodiversity goal and a global analysis of gaps in coverage of marine protected areas - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on which dataset was used. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___
<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

### 2016-03-04 ms_2_intro_res_disc.Rmd
updating refs, numbers etc in manuscript

``` {r calc species numbers, echo = FALSE}
n_spp_all    <- nrow(spp_all)
n_spp_am     <- nrow(spp_am)
n_spp_am_rev <- nrow(spp_am %>% filter(reviewed == 1))
n_spp_iucn   <- length(spp_iucn_maps$iucn_sid %>% unique())
n_spp_both   <- length(spp_map_pairs$iucn_sid %>% unique())

p_spp_am     <- round(nrow(spp_am)/n_spp_all * 100, 1)
p_spp_am_rev <- round(n_spp_am_rev/n_spp_am * 100, 1)
p_both_am    <- round(n_spp_both/n_spp_am   * 100, 1)
p_both_iucn  <- round(n_spp_both/n_spp_iucn * 100, 1)
p_both_all   <- round(n_spp_both/n_spp_all  * 100, 1)
```

@@

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) (REF).  These two spatial datasets have been used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015 _both_; Selig et al. 2013 _both_), evaluating global biodiversity patterns (Coll et al. 2010 _AM_; Martin et al. 2014 _both and others_, Pimm 2014 _IUCN, mostly focusing on terrestrial but some marine_), predicting range shifts (Molinos et al. 2015 _AquaMaps_, and setting conservation priorities (Klein et al. 2015 _AquaMaps_). <!-- BSH: Any of these use IUCN?  Would be good to make sure at least 2-3 examples were based on IUCN. ___none of these exclusively use IUCN - other suggestions?___-->

>>>>

@@

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) (IUCN 2015).  These two spatial datasets have been used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015 _both_; Selig et al. 2013 _both_), evaluating global biodiversity patterns (Coll et al. 2010 _AM_; Martin et al. 2014 _both and others_, Pimm 2014 _IUCN, mostly focusing on terrestrial but some marine_), predicting range shifts (Molinos et al. 2015 _AquaMaps_, and setting conservation priorities (Klein et al. 2015 _AquaMaps_). <!-- BSH: Any of these use IUCN?  Would be good to make sure at least 2-3 examples were based on IUCN. ___none of these exclusively use IUCN - other suggestions?___-->
